---
title: "Visualize data"
output: 
  html_notebook:
      toc: yes
      toc_float: yes
---

Graphs are often the starting point for statistical analysis.  One of the main advantages of **R** is how easy it is for the user to create many different kinds of graphs. We begin this chapter by studying conventional graphs, followed by an examination of some more complex representations. This final part uses the **ggplot2** package.

# 1. Conventional Graphical Functions

To begin with, it may be interesting to examine a few example of graphical representations
which can be constructed with **R**. We use the **demo** function:
```{r, eval=FALSE, echo=TRUE}
demo(graphics)
```

## The plot function

The **plot** function is a generic function used to represent all kinds of data. Classical use of the **plot** function consists of representing a scatterplot for a variable *y* according to another variable *x*. For example, to represent the graph of the function
$x\mapsto \sin(2\pi x)$ on $[0,1]$, at regular steps we use the
following commands:

```{r}
x <- seq(-2*pi,2*pi,by=0.05)
y <- sin(x)
plot(x,y) #dot representation (default)
plot(x,y,type="l") #line representation
```

We provide examples of representations for quantitative and qualitative variables. We use the data file **ozone.txt**, imported using

```{r}
path <- file.path("../DATA", "ozone.txt") 
ozone <- read.table(path)
summary(ozone)
```

Let us start by representing two quantitative variables: maximum ozone **maxO3** according to temperature **T12**:
```{r}
plot(ozone[,"T12"],ozone[,"maxO3"])
```

As the two variables are contained and named within the same table, a simpler syntax can be used, which automatically inserts the variables as labels for the axes:
```{r}
plot(maxO3~T12,data=ozone)
```
We can also use (more complicated)
```{r}
plot(ozone[,"T12"],ozone[,"maxO3"],xlab="T12",ylab="maxO3")
```
Functions **histogram**, **barplot** and **boxplot** allow to draw classical graphs:

```{r}
hist(ozone$maxO3,main="Histogram")
barplot(table(ozone$vent)/nrow(ozone),col="blue")
boxplot(maxO3~vent,data=ozone)
```

## Interactive graphs with rAmCharts

We can use this package to obtain dynamic graphs. It is easy, we just have to use the prefix **am** beforme the name of the function:

```{r message=FALSE, warning=FALSE}
library(rAmCharts)
amHist(ozone$maxO3)
amPlot(ozone,col=c("T9","T12"))
amBoxplot(maxO3~vent,data=ozone)
```


### Exercise 1

* Draw the sine function between 0 and $2\pi$.
* Add the following title: **plot of sine function**.


### Exercise 2

* Draw the pdf (probability distribution function) of the standard Gaussian distribution between $-4$ and 4 (use **dnorm**).
* Add a vertical dashed line of equation $x=0$ (use **abline**)
* On the same graph, draw Student's $t$-distribution to 5 and 30 degrees of freedom (use **dt**). Use the **lines** function
and a different colour for each line.
* Add a legend at the top left to differentiate between each distribution (use **legend**).

### Exercise 3 

We consider the **ozone** dataset. With the **layout** function, split the window into two lines with

  * the scatter plot **maxO3 vs T12** on the first line;
  * 2 graphs on the second lines: histogram of **T12** et boxplot of **maxO3**.



# 2. Ggplot2

ggplot2 is a plotting system for R based on the grammar of graphics (as **dplyr** to manipulate data). We can find documentation [here](http://ggplot2.org). We consider a subsample of the diamond dataset from the package **ggplot2** (or **tidyverse**):

```{r}
library(tidyverse)
set.seed(1234)
diamonds2 <- diamonds[sample(nrow(diamonds),5000),] 
summary(diamonds2)
help(diamonds)
```

Given a dataset, a graph is defined from many **layers**. We have to specify:

* the data
* the variables we want to plot
* the type of representation (scatterplot, boxplot...).

Ggplot graphs are defined from these layers. We indicate

* the data with **ggplot**
* the variables with **aes** (aesthetics)
* the type of representation with **geom_**

The scatterplot **carat vs price** is obtained with the **plot** function with
```{r}
plot(price~carat,data=diamonds2)
```

With **ggplot**, we use
```{r}
ggplot(diamonds2) #nothing
ggplot(diamonds2)+aes(x=carat,y=price) #nothing
ggplot(diamonds2)+aes(x=carat,y=price)+geom_point() #good
```


### Exercise 4

* Draw the histogram of **carat** (use **geom_histogram**)
* Draw the histogram of **carat** with 10 bins (**help(geom_histogram)**)
* Draw the barplot for the variable **cut** (use **geom_bar**)




## 2.1 ggplot grammar

In ggplot, the syntax is defined from independent elements. These elements define the **grammar** of **ggplot**. Main elements of the grammar include:

* **Data (ggplot)**: the dataset, it should be a dataframe
* **Aesthetics (aes)**: to describe the way that variables in the data are mapped. We have to specify **all** variables used to buid the graph in **aes**
* **Geometrics (geom_...)**: to control the type of plot
* **Statistics (stat_...)**: to describe transformation of the data
* **Scales (scale_...)**: to control the mapping from data to aesthetic attributes (change of colors...)

All these elements are conbined with a **+**.

### `Data and aesthetics`

These two elements specify the data and the variables we want to represent. For a scaterplot **price vs carat** we enter the command

```{r}
ggplot(diamonds2)+aes(x=carat,y=price)
```

**aes** also use arguments such as **color**, **size**, **fill**. We use these arguments as soon as a color or a size is defined from a **variable of the dataset**:

```{r eval=T}
ggplot(diamonds2)+aes(x=carat,y=price,color=cut)
```

### `Geometrics`

To obtain the graph, we need to precise the type of representation. We use **geometrics** to do that. For a scatter plot, we use **geom_point**:

```{r}
ggplot(diamonds2)+aes(x=carat,y=price,color=cut)+geom_point()
```

Observe that **ggplot** adds the lengend automatically. Exemples of **geometrics** are described here:

Geom   | Description| Aesthetics 
-------|------------|-----------
geom_point()| Scatter plot | x, y, shape, fill 
geom_line()|  Line (ordered according to x) | x, y, linetype
geom_abline()|  Line | slope, intercept 
geom_path() | Line (ordered according to the index) | x, y, linetype 
geom_text() | Text | x, y, label, hjust, vjust 
geom_rect() | Rectangle | xmin, xmax, ymin, ymax, fill, linetype
geom_polygon() | Polygone | x, y, fill, linetype
geom_segment() | Segment | x, y, fill, linetype 
geom_bar() | Barplot | x, fill, linetype, weight 
geom_histogram() | Histogram | x, fill, linetype, weight 
geom_boxplot() | Boxplots | x, y, fill, weight 
geom_density() | Density | x, y, fill, linetype 
geom_contour() | Contour lines | x, y, fill, linetype 
geom_smooth() | Smoothers (linear or non linear) | x, y, fill, linetype 
All | | color, size, group


### Exercise 5

* Draw the barplot of **cut** (with blue bars)
* Draw the barplot of **cut** with one color for each modality of cut.


### `Statistics` (this part can be omitted for beginners)

Many graphs need to transform the data to make the representation (barplot, histogram). Simple transformations can be obtained quickly. For instance we can draw the sine function with

```{r}
D <- data.frame(X=seq(-2*pi,2*pi,by=0.01))
ggplot(D)+aes(x=X,y=sin(X))+geom_line()
```

The sine transformation is precised in **aes**. For more complex transformations, we have to used **statistics**. A **stat** function takes a dataset as input and returns a dataset as output, and so a stat can add new variables to the original dataset. It is possible to map
aesthetics to these new variables. For example, stat_bin, the statistic used to make histograms, produces the following variables:

* `count`, the number of observations in each bin
* `density`, the density of observations in each bin (percentage of total / bar
width)
* `x`, the center of the bin

By default *geom_histogram* represents on the $y$-axis the number of observations in each bin (the outuput **count**). 

```{r}
ggplot(diamonds)+aes(x=price)+geom_histogram(bins=40)
```

For the density, we use
```{r}
ggplot(diamonds)+aes(x=price,y=..density..)+geom_histogram(bins=40)
```

*ggplot* propose another way to make the representations: we can use *stat_* instead of *geom_*. Formally, each stat function has a geom and each geom has a stat. For instance, 
```{r}
ggplot(diamonds2)+aes(x=carat,y=price)+geom_smooth(method="loess")
ggplot(diamonds2)+aes(x=carat,y=price)+stat_smooth(method="loess")
```
lead to the same graph. We can change the type of representation in the **stat_** with the argument **geom**:

```{r}
ggplot(diamonds2)+aes(x=carat,y=price)+stat_smooth(method="loess",geom="point")
```

Here are some examples of **stat functions**

 Stat   |  Description |  Parameters
--------|------------------|----------------
stat_identity() | No transformation |  
stat_bin() | Count | binwidth, origin 
stat_density() | Density | adjust, kernel 
stat_smooth() | Smoother | method, se 
stat_boxplot() | Boxplot | coef 

*stat* and *geom* are not always easy to combine. For beginners, we recommand to only use *geom*.

### Exercise 6 (optional)

We consider a color variable $X$ with probability distribution
$$P(X=red)=0.3,\ P(X=blue)=0.2,\ P(X=green)=0.4,\ P(X=black)=0.1$$
Draw the barplot of this distribution.


### `Scales`

Scales control the mapping from data to aesthetic attributes (change of colors, sizes...). We generally use this element at the end of the process to refine the graph. Scales are defined as follows:

* begin with **scale_**
* add the aesthetics we want to modify (**color**, **fill**, **x_**)
* end with the name of the scale (**manual**, **identity**...)

For instance,

```{r}
ggplot(diamonds2)+aes(x=carat,y=price,color=cut)+geom_point()+
scale_color_manual(values=c("Fair"="black","Good"="yellow",
"Very Good"="blue","Premium"="red","Ideal"="green"))
```

Here are the main scales:

aes | Discrete | Continuous
----|-----------|----------
Couleur (color et fill) | brewer | gradient 
  -  | grey | gradient2 
  -  | hue | gradientn 
 - | identity | 
 - | manual |
Position (x et y) | discrete | continous 
- | | date 
Forme | shape | 
- | identity | 
- | manual | 
Taille  | identity | size 
- | manual | 

Some examples:

* `color of a barplot`

```{r}
p1 <- ggplot(diamonds2)+aes(x=cut)+geom_bar(aes(fill=cut))
p1
```


We change colors by using the palette **Purples** :

```{r}
p1+scale_fill_brewer(palette="Purples")
```


* `Gradient color for a scatter plot` :
```{r}
p2 <- ggplot(diamonds2)+aes(x=carat,y=price)+geom_point(aes(color=depth))
p2
```

We change the gradient color 
```{r}
p2+scale_color_gradient(low="red",high="yellow")
```

*  `Change on the axis`
```{r}
p2+scale_x_continuous(breaks=seq(0.5,3,by=0.5))+scale_y_continuous(name="prix")+scale_color_gradient("Profondeur")
```


### `Group and facets`

**ggplot** allows to make representations for subgroup of individuals. We can proceed in two ways:

* to represent subgroup on the same graph, we use *group* in **aes**
* to represent subgroup on the different graphs, we use **facets**

We can represent (on the same graph) the smoother **price vs carat** for each modality of *cut* with 
```{r}
ggplot(diamonds2)+aes(x=carat,y=price,group=cut,color=cut)+geom_smooth(method="loess")
```

To obtain the representation on many graphs, we use

```{r}
ggplot(diamonds2)+aes(x=carat,y=price)+geom_smooth(method="loess")+facet_wrap(~cut)
ggplot(diamonds2)+aes(x=carat,y=price)+geom_smooth(method="loess")+facet_wrap(~cut,nrow=1)
```

*facet_grid* and *facet_wrap* do the same job but split the screen in different ways:

```{r}
ggplot(diamonds2)+aes(x=carat,y=price)+geom_point()+geom_smooth(method="lm")+facet_grid(color~cut)
ggplot(diamonds2)+aes(x=carat,y=price)+geom_point()+geom_smooth(method="lm")+facet_wrap(color~cut)
```


## 2.2 Complements

Syntax for **ggplot** is defined according to the following scheme:

ggplot()+aes()+geom_()+scale_()

It is really flexible: for instance **aes** could also be specified in **ggplot** or in **geom_**

```{r}
ggplot(diamonds2)+aes(x=carat,y=price)+geom_point()
ggplot(diamonds2,aes(x=carat,y=price))+geom_point()
ggplot(diamonds2)+geom_point(aes(x=carat,y=price))
```

We can also built a graph with many datasets:
```{r}
X <- seq(-2*pi,2*pi,by=0.001)
Y1 <- cos(X)
Y2 <- sin(X)
donnees1 <- data.frame(X,Y1)
donnees2 <- data.frame(X,Y2)
ggplot(donnees1)+geom_line(aes(x=X,y=Y1))+
geom_line(data=donnees2,aes(x=X,y=Y2),color="red")
```

Many other functions are proposed by **ggplot**:

* **ggtitle** to add a title
* **ggsave** ta save a graph
* **theme_** to change the theme of the graph

```{r}
p <- ggplot(diamonds2)+aes(x=carat,y=price,color=cut)+geom_point()
p+theme_bw()
p+theme_classic()
p+theme_grey()
p+theme_bw()
```


### Exercise 7

1. Draw the sine and cosine funtions on the same graph. You first use two datasets (one for the sine function, the other for the cosine function).


2. Do the same with one dataset and 2 geom_line


3. Do the same with one dataset and 1 geom_line. Hint: use the **gather** function of the package **tidyverse**.

4. Draw the two functions on two different graphs (use **facet_wrap**).


5. Do the same with the function **grid.arrange** from the package **gridExtra**.


### Exercise 8

We consider the dataset **mtcars**
```{r}
data(mtcars)
summary(mtcars)
```

1. Draw the histogram of **mpg** (use many number of bins)


2. Represent the density on the $y$-axis.


3. Draw the barplot of **cyl**.


4. Draw the scatter plot **disp vs mpg** for each value of **cyl** (one color for each value of **cyl**).


5. Add the linear smoother on each graph.



### Exercise 9

1. Draw the sine function on $[-2\pi,2\pi]$
2. Add the lines (in blue) of equation $y=1$ and $y=-1$. Use *size=2*.


### Exercise 10

1. Simulate a sample $(x_i,y_i),i=1,\dots,100$ according to the linear model
$$Y_i=3+X_i+\varepsilon_i$$
where $X_i$ are i.i.d. and uniform on $[0,1]$ and $\varepsilon_i$ are gaussian $N(0,0.2^2)$ (use **runif** and **rnorm**)


2. Draw the scatter plot **y vs x** and add the linear smoother.

3. Draw the residuals: add a vertical line from each point to the linear smoother (use **geom_segment**).





### Challenge (optional)

We consider the **diamonds** dataset.

1. Obtain the following graphs (use **coord_flip** for the second graph).

![](challenge1.pdf)
![](challenge2.pdf)
![](challenge3.pdf)



2. Add on the third graph a line for the quartiles of the variable **carat** (for each value of **cut**)


3. Draw the following graph (use the **ggstance** package).

![](challenge4.pdf)



### Exercise 11

We consider the dataset about the 4 tennis major tournaments in 2013 studied in the previous sheet.



1. Draw two histograms (one for Roland Garros, the other for wimbledon) to visualize the distribution of the number of aces per match.

2. Make a comparion of the number of aces per match between Roland Garros and Wimbledon with a boxplot.


3. Make a comparion of the net points attempted per match between Roland Garros and Wimbledon with a boxplot.


4. For the Roland Garros tournament, study with a boxplot the influence of the first serve percentage on the result of the match


5. Do the same for the Wimbledon Tournament.

6. Make a comparison of the first serve percentage of the winner of the match between Rolang Garros and Wimbledon.

