---
title: 'Tuto 2: R objects'
output: html_notebook
---

## Working directory

Before starting, you have to create a directory where you put all the files of the tutorial. The command

```{r}
getwd()
```
indicates the directory where **R** works. The command **setwd** allows to change this directory. You can also use the menu **Session -> Set Working directory -> Choose directory...**

## Creating objects

### Numeric
An object can be created by being assigned to one of the three operators **<-**, **->**, **=** and by naming 

```{r}
b<-41.3  # set up object b consisting of one number: 41.3 
x<-b     # b is assigned to x
x=b      # b is assigned to x
b->x     # b is assigned to x
is.numeric(b)
mode(b)
```

### Character

```{r}
x <- "La mort"
y <- "aux trousses"
paste(x,y)
is.character(x)
```

### Logical (Boolean)

```{r}
x <- TRUE
is.logical(x)
mode(x)
a <- 1
a==1
a!=1
a<0
a>0
```

### Vector

There are different ways to define a vector:

* collect function **c**
```{r}
x <- c(1.2,5,9,11)
x
```

* sequence operator **:**

```{r}
1:5
```

*  sequence function **seq**

```{r}
seq(1,10,by=2)
seq(0,1,length=10)
```

* The **rep** function

```{r}
rep(1,4)
rep(c(1,3),each=3)
```

We can also create *character* or *logical* vectors

```{r}
x <- c("A","B","C")
x <- rep("A",5)
paste("X",1:5,sep="-")
substr("statistician",5,9)
```


### Selecting part of a vector

Selections are made using the selection operator **[ ]**

```{r}
x <- c(-4,-3,1,3,5,8,0)
x[2]
x[c(2,5)]
x>0
x[x>0]
```


### Operations

Explain the following commands
```{r}
x <- seq(-10,10,by=2)
y <- 1:length(x)
x+y
x*y
z <- x>0
x*z
```

## Factors

Factors are particular vectors very useful when we deal with categorical variables.

```{r}
V1 <- factor(c("less20years","more50years","less20years","more50years","less20years"))
V1
levels(V1)
levels(V1) <- c("Young","Old")
V1
```


### Exercise 1

1. Compute the mean, the sum, the median and the variance of the vector (1,3,8,9,11)


2. Create the following vectors using the **rep** function:

vec1 = 1 2 3 4 5 1 2 3 4 5 1 2 3 4 5 

vec2 = 1 1 1 2 2 2 3 3 3 4 4 4 5 5 5

vec3 = 1 1 2 2 2 3 3 3 3 4 4 4 4 4


3. Create the following vector using the **paste** function:

vec4 = "A0)" "A1)" "A2)" "A3)" "A4)" "A5)" "A6)" "A7)" "A8)" "A9)" "A10)"


4. **letters** is a vector containing the 26 letters of the alphabet. Find the index of the letter $q$ and create the vector containing "a1","b2",$\dots$ until $q$ with the index of $q$.




## Matrix

The **matrix** function allows to define matrices
```{r}
m <- matrix(1:4,ncol=2)
m
m <- matrix(1:4,nrow=2)
m
m <- matrix(1:4,nrow=2,byrow=TRUE)
dim(m)
```

The position of an element within a matrix is indicated by its row and column number. Thus, to select the term in the second row and the first column, we execute
```{r}
m[2,1]
```

We can also extract rows or columns of matrices:
```{r}
m[1,] #first row
m[,2] #second column
```

It is easy to make classical operations on matrices

```{r}
det(m) #determinant
solve(m) #inverse
t(m) #transpose
n <- matrix(5:8,nrow=2)
m+n
m*n #be carefull: Hadamart product
m%*%n #matrix product
eigen(m) #eigen decomposition
```

## Lists

A list is a heterogeneous object.  It is a set of ranked objects which do not always have the same mode or length. The objects are referred to as components of the list. For instance, we can put a vector and a matrix in a list with

```{r}
mylist <- list(vector=rep(1:5),mat=matrix(1:8,nrow=2))
mylist
length(mylist)
```
To extract a component from the list, we can simply indicate the position of the element that we want to extract. **[[  ]]** are used to identify the element in the list:

```{r}
mylist[[1]]
```

We can also use the name of the element, if it has one, which can be written in two ways:
```{r}
mylist$mat
mylist[["mat"]]
```


## Dataframe

Dataframes are special lists with components of the same length but with potentially different modes. The data tables generally used in statistics are often referred to as dataframe. Indeed, a data table is made up of quantitative and/or qualitative variables taken from the same individuals.

```{r}
name <- c("Paul","Mary","Steven","Charlotte","Peter")
sex <- factor(c("M","F","M","F","M"))
size <- c(180,165,168,170,175)
data <- data.frame(name,sex,size)
summary(data)
```

Observe that *name* is character vector, *sex* a factor and *size* a numeric vector.

Extraction is similar to matrix and list

```{r}
data[2,3]
data[,2]
data$sex
```


## Some basic (and important) functions

* **summary** produces result summaries of an object

```{r}
summary(data)
summary(1:10)
```

* mean, sum, median, var, min, max... (easy to understand)

* *sort*, *order*

```{r}
x <- c(1,8,5,4)
sort(x)
order(x)
```

* **apply** apply a function **f** to the rows or columns of a matrix or a dataframe

```{r}
V1 <- 1:10
V2 <- seq(-20,25,length=10)
df <- data.frame(V1,V2)
apply(df,1,mean)
apply(df,2,sum)
```

### Exercise 2

1. Create the following matrix *mat* (with the column and row names, you can use **rownames** and **colnames** functions):

|      |column 1 | column 2 |column 3 | column 4 |
|:----:|:-------:|:--------:|:-------:|:--------:|
|row-1 |    1    |     5    |    5    |    0     |
|row-2 |    0    |     5    |    6    |    1     |
|row-3 |    3    |     0    |    3    |    3     |
|row-4 |    4    |     4    |    4    |    2     |


2. Create a vector containing the diagonal elements of the matrix **mat**.

3. Create a matrix containing the first 2 rows of **mat**.

4. Create a matrix containing the last 2 columns of **mat**.

5. Calculate the determinant and then invert the matrix using the appropriate functions.

### Exercise 3

We consider the **iris** dataset available in **R** (use **data(iris)** to load it and then **head(iris)** to visualise the first rows).

```{r}
data(iris)
head(iris)
```


1. Compute the mean and the variance for the variables **Sepal.Width** and **Petal.Length** :


2. Create a sub-dataset comprising only the data for the category *versicolor* of the variable *species* (call this new dataset **iris2**).


3. Sort the data in **iris2** in descending order according to the variable **Sepal.Length** (you can use the **order** function).


4. Compute the mean of **Sepal.Length** for each species.


5. Add a variable (called **sum.Petal**) in the dataframe iris which corresponds to the sum of **Petal.Length** and **Petal.Width**.



### Exercise 4

1. Calculate the benchmark statistics (mean, min, max, etc.) for the three variables of the **ethanol** dataset (available in the *lattice* package). 


2. Calculate the quartiles for each of the three variables. To do this, use the **apply** function with the **quantile** function.


3. Again with the **apply** function, calculate all the deciles for each of the three variables using the argument **probs** of the **quantile** function.



### Exercise 5

We consider the **predidents** dataset.

```{r}
data("presidents")
df <- matrix(presidents,ncol=4,byrow=T)
```

1. Is there at least one missing value in the 20th line of the dataset (you can use the **any** function).


2. Which lines in the table contain at least one missing data? You can use **which**.

3. Delete the rows of the table that contain at least one missing data.


