---
title: "Mapping with leaflet"
output: 
  html_notebook:
#    toc: true
#    toc_float: true
---



Leaflet is a R package which allows to produce interactive maps (you can find information [here](https://rstudio.github.io/leaflet/)). We can obtain a background map with

```{r}
library(leaflet)
m <- leaflet()
m <- addTiles(m)
m
```

It is possible to use the pipe operator of *dplyr*
```{r message=FALSE, warning=FALSE}
library(tidyverse)
leaflet()%>%addTiles()
```

Many formats of background maps are available (some examples [here](http://leaflet-extras.github.io/leaflet-providers/preview/)). For instance

```{r}
Paris <- c(2.35222,48.856614)
m2 <- leaflet() %>% setView(lng = Paris[1], lat = Paris[2], zoom = 12) %>% addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png")
m2 %>% addProviderTiles("Stamen.Toner")
m2 %>% addProviderTiles("Stamen.Toner")
m2%>% addProviderTiles("OpenStreetMap.BlackAndWhite")
m2%>% addProviderTiles("Thunderforest.Transport")
```

**addMarkers** or **addCircles** can be used to add markers:
```{r}
data(quakes)
leaflet(data = quakes[1:20,]) %>% addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
  addMarkers(~long, ~lat, popup = ~as.character(mag))
```

Observe that we have to use the tilda character **~** when we want to use the names of the dataframe to obtain the map. 

Popups can add informations:

```{r}
content <- paste(sep = "<br/>",
  "<b><a href='http://www.samurainoodle.com'>Samurai Noodle</a></b>",
  "606 5th Ave. S",
  "Seattle, WA 98138"
)

leaflet() %>% addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
  addPopups(-122.327298, 47.597131, content,
    options = popupOptions(closeButton = FALSE)
  )
```


### Exercise 1

Put a popup which allows to localize **Edhec Business School**. Add on this popup the website of the school. You can obtain longitude and latitude of Edhec at the following [url](https://www.coordonnees-gps.fr)



### Exercise 2: city bike - Paris

1. Import this [dataset](https://opendata.paris.fr/explore/dataset/velib-disponibilite-en-temps-reel/export/)


2. Explain the variables

3. Suppress lines without longitude and latitude.

4. Describe the variable **geo**

5. Create one variable **lon** for the longitude of the station a one variable **lat** for the latitude.

6. Represent the stations on a leaflet background map.

7. Add a popup which indicates the number of available bikes (electric+mecanic) when we click on the station (you can use the **popup** option in the function **addCircleMarkers**). You fist have to define a variable **numBikesAvailable** which corresponds to the sum of **Nombre.de.vélo.mécanique** and **Nombre.vélo.électrique**.

8. Add the name (**Nom.de.la.station**) of the station in the popup.

9. Make points of different sizes in terms of the proportion of available bikes (**= numBikesAvailable/ Nombres.de.bornes.en.station**).
