---
title: 'Tuto 1: RStudio environment'
output: html_notebook
---



**R** is a free software for statistical computing and graphics. It is freely distributed by **CRAN** (Comprehensive R Archive Network) at the following adress: [https://www.r-project.org](https://www.r-project.org). The installation varies according to the operating system (Windows, Mac OS, Linux). But it is easy, you just have to follow the instructions.


**RStudio** is an integrated development environment (IDE) for R. It includes a console, syntax-highlighting editor that supports direct code execution, as well as tools for plotting, history, debugging and workspace management. It is also freely distributed at the address [Rstudio](https://www.rstudio.com). 

The screen is divided into 4 windows:

* *Console*: where you can type command and see output (enter 1+2)
* *Workspace an History*: show the active object (enter a <- 1+2)
* *Files Plots...*: show all files anf folders in the workspace, see output graph, install packages...
* *R script*: where you keep a record of your work. Don't forget to regularly save this files!

## R Script

There are different ways to work with Rstudio. Basically, you 

* open a **script** 
* enter the command in the script
* see the output in the console (click on the run icon)
* save the script

## Notebook and Rmarkdown

Recently (two years ago), a notebook has been developed to produce high quality reports in html or pdf formats. An R Notebook is an R Markdown document with chunks that can be executed independently and interactively, with output visible immediately beneath the input. You can create a new notebook in RStudio with the menu command *File -> New File -> R Notebook*. Rcode are embedded into chunks: menu *Insert -> R* or *Ctrl + Alt + I*

The [R markwdown cheat Sheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf) describes the syntax for Markdwon and notebook: tabular, graphics, equation with latex (you need to install *latex* to see the equations)...
$$\int f(x)\ dx=1$$



Here are some classical commands:

* *italic* and **bold**
* unordrerd list:
    - item 1
    - item 2
* ordered list:
    1. item 1
    2. item 2
* table:

| Col1 | Col2 | Col3 |
|------|------|------|
Row1 | 1 | 2| 3 |
Row2 | 1 | 2| 3 |

**R commands** need to be insert in chunks. You can quickly insert chunks like these into your file with

* the keyboard shortcut Ctrl + Alt + I (OS X: Cmd + Option + I)
* the Add Chunk  command in the editor toolbar
* or by typing the chunk delimiters ```{r} and ```.

You just have to click on *"Preview"* to see the html output file.


### Exercise

1. Open a new R Notebook file (*File -> New File -> R Notebook*)
2. Create a section named **Cosine**
3. Draw the cosine function: write the following code in a chunk


```{r,echo=TRUE,eval=FALSE}
x <- seq(-2*pi,2*pi,by=0.01)
y <- cos(x)
plot(x,y,type="l")
```

4. Execute the chunk (click on the green triangle).
5. Click on the icon *Preview*.
6. Add a second section *Sine* where you draw the sine function.



The source **Rmd** files can be obtained from the html file: just click on *Code -> Download Rmd*. From now on, we **always** work on notebook.

## R presentation

**Rstudio** also provides many environments to produce slides. You can create a presentation in RStudio with the menu command *File -> New File -> R Markdown -> Presentation*. Then we have to select *ioslides* or *slidy*.  You have to use the same syntax as for Notebook. Separation between slides are made with ## and R commands are always in chunk.


### Exercice

Create 2 slides:

1. Title: **cosine** and draw the cosine function.
2. Title: **sine** and draw the sine function.

## Packages

A package is simply a set of **R** programs which supplements and enhances the functions of **R**. Packages are generally reserved for specific methods or fields of applications. There are more than 13 000 packages available at the url [https://cran.r-project.org](https://cran.r-project.org). You can install a package

* with the function **install.packages** (in the console)
* by clicking on the *Packages* icons.

Once the package has been installed, you can call it with the **library** function:

```{r, eval=FALSE, include=TRUE}
install.packages(package.name)
library(packages.name)
```


### Exercise

1. Execute

```{r,eval=FALSE,include=TRUE}
iris %>% summarize(mean_Petal=mean(Petal.Length))
```

What happens?

2. Install and load the package **tidyverse** and re-execute the previous code.

```{r,eval=FALSE,include=TRUE}
install.packages("tidyverse")
```


```{r}
library(tidyverse)
iris %>% summarize(mean_Petal=mean(Petal.Length))
```



